---
title: "Comparing Rent Prices in SoCal Housing and Making Inferences"
editor: visual
---

## Motivation and Context

Southern California has a lot of reasons for people to live there: Los Angeles is full of many areas to tour, though it also has a high crime rate. Meanwhile, Orange County is home to Disneyland, which is labeled as "the happiest place on earth". Both counties also are home to beaches that attract tourists that have to travel and spend money doing so as opposed to people that live locally who can just drive to those beaches. My objective is to see if I can take any insight from house data in Southern California in the year 2025 to see if there is a difference between the prices in each county and do an inference on it.

```{r}
#| label: import relevant datasets
#| echo: false
#| message: false

SoCalRent2025 <- readr::read_csv(here::here("../Data", "2025SoCalRent - Sheet1.csv"))

```

## Main Objective

The objective is to compare rent prices in Los Angeles County and Orange County. My goal is to see the difference between the two  I am accomplishing this by doing some data wrangling to remove some prices that are clear outliers before doing analysis on the rest of the data. We are then going to make an inference about the relationship between price and county using Bayesian Statistics. We collected data on 7 variables: `baths` means number of bathrooms in the house while `beds` means the number of bedrooms in the house.

```{r}
#| label: libraries

library(tidyverse)
library(janitor)
library(httr2)
library(robotstxt)
library(rstanarm)
```

## Design and Data Collection

The data was collected from somebody looking on Redfin for houses (credit to Natalie): it shows the city, price, beds, baths, area, type, and county that the data was collected in. This was collected at a specific date in 2025 (April 15), however, which very well could limit the analysis as it may not reflect things happening a month later when this project is finished. The variables also aren't the best at giving insight as to why there may be a difference in prices between the cities because they only describe the house, not the other variables of the city.

## Data Massaging

Something I noticed about the data was that SoCalRent2025 has commas and dashes in some of the observations of square feet. The variable sq ft is a character variable, so I had to make it numeric, but I had to change some things. First, I removed the commas from the numbers in the dataset to make them ready to be converted to a numeric variable.

```{r}
#| label: Remove commas in all numbers

SoCalRent2025 <- SoCalRent2025 %>% 
  mutate(
    `sq ft` = str_remove(`sq ft`, ",") #Remove commas in all numbers that have them
  )
```

That would make everything ready for converting square feet into a numerical variable, except that some of the entries in `sq ft` are what appear to be ranges with a "-" character. To do this, I ultimately decided to make two new variables: a minimum and a maximum variable for square feet so that both are represented to look at additional nuances of area.

```{r}
#| label: two different datasets that base off the range

SoCalRent2025 <- SoCalRent2025 %>% 
  mutate(
    min_sq_ft = if_else(str_detect(`sq ft`, "-"), 
    true = as.numeric(str_extract(`sq ft`, "[0-9]+")),
    false = as.numeric(`sq ft`),
    missing = NA), # Code to put the minimum square feet of the house, if there is a range of areas
    max_sq_ft = if_else(str_detect(`sq ft`, "-"),
    true = as.numeric(str_remove(str_extract(`sq ft`, "-[0-9]+$"), "-")),
    false = as.numeric(`sq ft`),
    missing = NA) # Code to put the maximum square feet of the house, if there is a range of areas
    )

# If the house has no range, the minimum and maximum are the same.

```

## Exploratory Data Analysis

Let's start with a general summary

```{r}
#| label: exploring the data

SoCalRent2025 |>
  group_by(county) %>% 
  summarize(
    mean_price = mean(price),
    n = n() # Total number of houses by county
  )
  
```

This summary shows that the distribution is fairly even, with each county having 100 houses apiece. This was created ensure everything is almost equally represented. The dataset SoCalRent2025 features three counties: Los Angeles (LA), Orange (OC), and Riverside (RIV). The main thing that stood out was the mean price of LA County and Orange County are both much higher than that of Riverside, though I am mainly interested in focusing on LA County and Orange County to see if there was a difference between those two.

```{r}
#| label: seeing the range of prices for each type of house

ggplot(SoCalRent2025, aes(x = price, y = county)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "County"
  ) # Boxplots to see the relationship between price and county

 SoCalRent2025 %>%
   group_by(county) %>%
     summarize(
       min_price = min(price),
       q1_price = quantile(price, 0.25),
       med_price = median(price),
       q3_price = quantile(price, 0.75),
       max_price = max(price)
     ) # Numerical summary of the boxplot
```

This graph is displaying the amount of outliers that are present in the dataset in terms of price, and there are two are stick out like a sore thumb. Both are prices at or above \$20000. The table is a numerical summary of the boxplots, which also shows the distribution of prices and shows how Orange County tends to have more expensive houses than the others, at least in this dataset. Both were created to show the distribution of prices by county.

```{r}
#| label: investigating the outliers in terms of price

SoCalRent2025 %>%
   filter(price >= 20000)
```

This figure shows the two outliers that I noticed. Right off the bat, a rent for \$30000 in OC, a house in Anaheim that has 2 bedrooms and 2 bathrooms. It doesn't sound like it is the norm, so we will remove this outlier. We can also filter out the \$20000 rent in LA county because that is the other thing that is way above the rest of the prices, plus it's a West Hills house that has the highest square feet, bedrooms, and bathrooms of all the remaining houses.

```{r}
#| label: new price by county summary and price five-number summary

SoCalRent2025_1 <- SoCalRent2025 %>% 
  filter(price < 20000) # creating SoCalRent2025_1

ggplot(SoCalRent2025_1, aes(x = price, y = county)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "County"
  )

SoCalRent2025_1 %>%
  group_by(county) %>%
  summarize(
    mean_price = mean(price)
  ) # numerical summary of the boxplot

```

These figures show the summaries after we have filtered out those data points. In the first figure, we bring it down to a maximum price of \$14000 and a median price of \$3097.5, which, considering how many houses over \$10000 are in the dataset, make sense to keep, even if they're considered outliers, since they're close enough to the rest of the data. The second figure shows that Orange County, on average, has higher rent than Los Angeles County. One variable in the dataset that makes sense to explain this is square feet of the houses:

```{r}
#| label: boxplots for square feet by county

ggplot(SoCalRent2025_1, aes(x = min_sq_ft, y = county)) +
   geom_boxplot()

ggplot(SoCalRent2025_1, aes(x = max_sq_ft, y = county)) +
   geom_boxplot()
```

This histogram shows the square feet by county, adjusted by the minimum and maximum square feet for each house that had a range of areas. It was created to display the distribution of square feet by county as part of our exploratory data analysis.

```{r}
#| label: show the relationship between type of house and square feet

ggplot(SoCalRent2025_1, aes(x = min_sq_ft, y = type)) +
   geom_boxplot() # using minimum square feet values

ggplot(SoCalRent2025_1, aes(x = max_sq_ft, y = type)) +
   geom_boxplot() # using maximum square feet values
```

This boxplot was created as a potential variable to explain square feet. We see from this graph that houses tend to be bigger than both condos and apartments. It is worth noting that apartments were the only things that had a range of values for their square feet, so not much changed between the condos and houses, but it is interesting that when we take the maximum square feet value for each apartment, they are very similar to condos.

```{r}
#| label: show each type of house by county

SoCalRent2025_1 %>% 
  group_by(county, type) %>% 
  summarize(
    count = n()
  )
```

Now we have to explore the distribution of house type by county, and it appears very even except for Riverside only having 25 condos. This ties into the counties all having similar average square feet in their houses, whether they were using the minimums or maximums.

Now, there are two different values due to some having ranges as part of their dataset. Here is a summary of the boxplots:

```{r}
#| label: updated summary of the minimum square feet values

SoCalRent2025_1 %>%
  filter(!is.na(min_sq_ft)) |>
   group_by(county) %>% 
     summarize(
       minimum_sq_ft = min(min_sq_ft),
       q1_sq_ft = quantile(min_sq_ft, 0.25),
       med_sq_ft = median(min_sq_ft),
       q3_sq_ft = quantile(min_sq_ft, 0.75),
       max_sq_ft = max(min_sq_ft),
       mean_sq_ft = mean(min_sq_ft)
     ) # five-number summary with minimum square feet values
```

This first one is the square feet when taking all the minimum values for those in the range. We see there are places where they are very similar, such as the median, though the mean square feet of a residence in Orange County is relatively high compared to the mean square feet of the other two when using minimum values.

```{r}
#| label: updated summary of the maximum square feet values

SoCalRent2025_1 %>%
  filter(!is.na(max_sq_ft)) %>% 
   group_by(county) %>%
     summarize(
       min_sq_ft = min(max_sq_ft),
       q1_sq_ft = quantile(max_sq_ft, 0.25),
       med_sq_ft = median(max_sq_ft),
       q3_sq_ft = quantile(max_sq_ft, 0.75),
       maximum_sq_ft = max(max_sq_ft),
       mean_sq_ft = mean(max_sq_ft)
     ) # five-number summary with maximum square feet values
```

So the datasets have quite some notable differences, though it may not be a fair representation as some of the minimum values by county don't even change, labeling things as outliers when we look at all maximum prices even though they are not when we look at all the minimum prices. Square feet doesn't appear to play a big factor, as they are all extremely similar between the counties.

And now, a histogram that shows the prices between the counties:

```{r}
#| label: histogram of final dataset by price

condition <- if_else(SoCalRent2025_1$price < 6000, "True", "False") # colors will be different depending on if price < 6000 or not on a histogram

ggplot(SoCalRent2025_1, aes(x = price, fill = condition)) +
  geom_histogram(bins = 15) +
  labs(
    title = "Rent Distrbution by County",
    x = "Price",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    plot.margin = margin(t = 10, r = 30, b = 30, l = 10),
    axis.title.x = element_text( # X-axis label
      size = 20
    ),
    axis.title.y = element_text( # X-axis label
      size = 20
    ),
    axis.text.x = element_text(
      size = 12
    ),
    axis.text.y = element_text(
      size = 16
    ),
    panel.spacing.x = unit(20, "points"),
  ) +
  facet_wrap(vars(county), ncol = 3, labeller = labeller(
               county = c(
                 "LA" = "Los Angeles County",
                 "OC" = "Orange County",
                 "RIV" = "Riverside County"
               )
             )) + # plots for each county
  scale_fill_manual(values = c("True" = "red", "False" = "blue"), labels = c(">=6000", "<6000")) # legend to show the condition
```

We can see that both Los Angeles County and Orange County have houses at or above \$6000 in rent while none in Riverside have higher than about \$5000. This definitely shows that Riverside houses tend to be cheaper to rent than those in Los Angeles County and Orange County. However, what I am mainly interested in is what the relationship between Los Angeles County and Orange County is like.

## Modeling

In order to see the true relationship, let's do some Bayesian Inference. This allows us to use data we have gathered and explored while incorporating it into a model.

```{r}
#| label: fit Bayesian regression model

bayes_SoCalRent <- stan_glm(
  price ~ county,
  data = SoCalRent2025_1,
  family = "gaussian",
  prior = normal(location = 1000, autoscale = TRUE),
  prior_intercept = default_prior_intercept(),
  seed = 1
)

SoCalRent_dist <- bayes_SoCalRent %>% 
  as.data.frame() # making the posterior distribution

head(SoCalRent_dist)
```
I chose the prior of a normal distribution centered around 1000. I think that there is some difference between Orange County and Los Angeles County in terms of rent prices and that Orange County houses tend to be higher than Los Angeles County houses. This makes a posterior distribution, which reflects updated beliefs after we have seen the housing rent data. This gives us an idea of what the relationship between the counties looks like.

The reference level is Los Angeles County, which means that Orange County and Riverside County will appear in the table as their respective variables, but not Los Angeles County since we are comparing both Orange County and Riverside County to Los Angeles County.

```{r}
#| label: histogram of the posterior distribution

plot(
  bayes_SoCalRent,
  pars = c("countyOC", "countyRIV"),
  plotfun = "hist",
  bins = 15
) +
  labs(x = "Slope by county")
```

We can see from the graph that the rent in LA is clearly higher than in Riverside, but there doesn't seem to be a significant difference between the rents of LA and Orange County.

```{r}
#| label: create posterior interval

bayes_SoCalRent |>
  posterior_interval(
      pars = c("countyOC", "countyRIV"),
      prob = 0.90
    )
```

This interval is a numerical summary of the histogram above. We can see that there is a 90% chance that the slope of the population model, assuming everything else stays constant, for an Orange County house compared to one in LA is between -76.47405 and 841.7148. This implies that there is no distinct difference in prices between the counties of LA and OC. There is also a 90% chance that the slope of the population model, assuming everything else stays constant, for a Riverside house compared to one in LA is between -2090.49632 and -1140.3071, meaning that Riverside houses tend to be cheaper to rent than LA houses. These values show that it is pretty unlikely that the average price of a house in LA is lower than one in Riverside, as both ends of the interval are negative, meaning that all positive values are towards the right end of the tail.

Doing Bayesian Inference allows us to make statements about our model relating county and price, with an estimated probability of 90% that the true model sees no relationship between the county a house is in and its rent price between the counties of LA and Orange when every other variable is the same. Between LA County and Riverside County, however, there is an approximately 90% chance that the true model has houses in LA County at higher prices than Riverside County should every other variable be the same.





## Conclusion

Los Angeles County and Orange County are both great places to live in, and the rent of the houses in those places are very similar. There doesn't appear to be a significant difference between the two counties. Riverside houses are much cheaper, but they also don't offer the same things both Los Angeles County and Orange County do.




## Limitations

The data was collected on a specific date, so things may have changed by then. The problem is since they can change on a day-by-day basis, the data is not always the most accurate since houses can be taken off or rent could go up. Also, as mentioned earlier, there isn't variables that are about the city, as the only one I could think of that was in the dataset with any real impact was square feet, and I didn't feel like there was much of a difference between the counties' average square feet.
