---
title: "Housing"
editor: visual
---

```{r}
#| label: libraries

library(tidyverse)
library(janitor)
library(httr2)
library(robotstxt)
```

```{r}
#| label: import-relevant-datasets
#| fig-cap: Code to import relevant datasets

SoCalRent2025 <- readr::read_csv(here::here("../Data", "2025SoCalRent - Sheet1.csv"))

SoCalRent2025 <- SoCalRent2025 %>% 
  mutate(
    `sq ft` = str_remove(`sq ft`, ",")
  )
```

```{r}
#| label: exploring the data

SoCalRent2025 |>
  group_by(county) %>% 
  summarize(
    mean_price = mean(price),
    n = n()
  )
  
```

The dataset SoCalRent2025 features three counties: Los Angeles (LA), Orange (OC), and Riverside (RIV). The distribution is fairly even, with each county having 100 houses apiece. We shall split it into three different datasets:

```{r}
#| label: make separate datasets to do separate explanatory analysis on each county

SoCalRent2025_LAC <- SoCalRent2025 %>% 
  filter(county == "LA")

SoCalRent2025_OC <- SoCalRent2025 %>% 
  filter(county == "OC")

SoCalRent2025_RIV <- SoCalRent2025 %>% 
  filter(county == "RIV")
```

As I was doing my exploratory data analysis, I realized that SoCalRent2025 has commas and dashes in some of the observations of square feet. The variable sq ft is a character variable, so I had to make it numeric, but I had to change some things. First, I removed the commas from the numbers in @import-relevant-datasets to make them ready to be converted to a numeric variable. Then, I noticed a bunch of houses that are a range. I decided to make two separate datasets, one with the minimum values and the other with the maximum values. This is to cover the extremities.

```{r}
#| label: two different datasets that base off the range
SoCalRent2025_minimums <- SoCalRent2025 %>%
  filter(!is.na(`sq ft`)) %>% 
  mutate(
    `sq ft` = as.numeric(str_extract(`sq ft`, "[0-9]+"))
  )

SoCalRent2025_maximums <- SoCalRent2025 %>% 
  filter(!is.na(`sq ft`)) %>% 
  mutate(
    `sq ft` = as.numeric(if_else(str_detect(`sq ft`, "-"),
      true = str_remove(str_extract(`sq ft`, "-[0-9]+$"), "-"),
      false = str_extract(`sq ft`, "[0-9]+"))
      )
  )
```

Another thing that stood out was the mean price of LA County and Orange County are both much higher than that of Riverside, and I mainly wanted to compare those two counties, so I want to see if the housing prices are significantly different, and potentially why they are if that's the case.

```{r}
#| label: seeing the range of prices for each type of house

ggplot(SoCalRent2025, aes(x = price, y = county)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "County"
  )

 SoCalRent2025 %>%
   group_by(county) %>%
     summarize(
       min_price = min(price),
       q1_price = quantile(price, 0.25),
       med_price = median(price),
       q3_price = quantile(price, 0.75),
       max_price = max(price)
     )
 
 SoCalRent2025 %>%
   filter(price >= 20000)
```

Right off the bat, a rent for \$30000 in OC, a house in Anaheim that has 2 bedrooms and 2 bathrooms. It doesn't sound like it is the norm, so we will remove this outlier. We can also filter out the \$20000 rent in LA county, though that one has 5 bedrooms and 4 bathrooms, among the highest amount of either within the entire dataset, so I am content with keeping it there for now.

```{r}
#| label: removing the massive outliers in terms of Price and Square Feet

SoCalRent2025_1 <- SoCalRent2025 %>% 
  filter(price < 30000) %>% 
  mutate(
    `sq ft` = as.numeric(`sq ft`)
  )

SoCalRent2025_1 %>%
  group_by(county) %>%
  summarize(
    mean_price = mean(price)
  )

SoCalRent2025_1 %>%
  summarize(
    min_price = min(price),
    q1_price = quantile(price, 0.25),
    med_price = median(price),
    q3_price = quantile(price, 0.75),
    max_price = max(price)
  )

ggplot(SoCalRent2025_1, aes(x = `sq ft`, y = type)) +
   geom_boxplot()

SoCalRent2025_1 %>% 
  filter(`sq ft` > 4000)

# ggplot(LaRent_1, aes(x = SqFt, y = Type)) +
#   geom_boxplot()
# 
# LaRent_1 %>% 
#   summarize(
#     min_sqft = min(SqFt),
#     q1_sqft = quantile(SqFt, 0.25),
#     med_sqft = median(SqFt),
#     q3_sqft = quantile(SqFt, 0.75),
#     max_sqft = max(SqFt)
#   )
```

When we remove it, the average rent price in orange county is now \$4557.73, cutting the difference between the mean OC rent price with the outlier and the mean LA rent price, but it's still pretty high. It also appears that the house with the second-highest price, which is in West Hills, also has the highest remaining price, square feet, bedrooms, and bathrooms of all the houses, so I'll remove it.

```{r}
#| label: updated plots

SoCalRent2025_1 <- SoCalRent2025_1 %>% 
  filter(`sq ft` < 4000)

SoCalRent2025_1 %>%
   group_by(county) %>%
     summarize(
       min_price = min(price),
       q1_price = quantile(price, 0.25),
       med_price = median(price),
       q3_price = quantile(price, 0.75),
       max_price = max(price)
     )

ggplot(SoCalRent2025_1, aes(x = price, y = county)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "County"
  )

ggplot(SoCalRent2025_1, aes(x = price, y = type)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "Type"
  )

SoCalRent2025_1 %>% 
  filter(price >= 12500)

SoCalRent2025_1 %>% 
  filter(`sq ft` >= 3000)
```

The boxplots don't seem to be too bad, but looking at the outliers by county in terms of price reveals a condo that can be rented for \$14000. I don't feel like that should be there

```{r}

```
