---
title: "Comparing Rent Prices in SoCal Housing and Making Inferences"
editor: visual
---

## Motivation and Context

Southern California has a lot of reasons for people to live there: Los Angeles is full of many areas to tour, though it also has a high crime rate. Meanwhile, Orange County is home to Disneyland, which is labeled as "the happiest place on earth". Both counties also are home to beaches that attract tourists that have to travel and spend money doing so as opposed to people that live locally who can just drive to those beaches. My objective is to see if I can take any insight from house data in Southern California in the year 2025 to see if there is a difference between the prices in each county and do an inference on it.

```{r}
#| label: import relevant datasets
#| echo: false
#| message: false

SoCalRent2025 <- readr::read_csv(here::here("../Data", "2025SoCalRent - Sheet1.csv"))

```

## Main Objective

The objective is to do some data wrangling to remove some prices that are clear outliers before doing analysis on the rest of the data. We are then going to make an inference about the relationship between price and county using Bayesian Statistics. We collected data on 7 variables: `baths` means number of bathrooms in the house while `beds` means the number of bedrooms in the house.

```{r}
#| label: libraries

library(tidyverse)
library(janitor)
library(httr2)
library(robotstxt)
library(rstanarm)
```

## Design and Data Collection

The data was collected from somebody looking on Zillow for houses: it shows the city, price, beds, baths, area, type, and county that the data was collected in. This was collected at a specific date in 2025 (April 15), however, which very well could limit the analysis as it may not reflect things happening a month later when this project is finished. The variables also aren't the best at giving insight as to why there may be a difference in prices between the cities because they only describe the house, not the other variables of the city.

## Data Massaging

Something I noticed about the data was that SoCalRent2025 has commas and dashes in some of the observations of square feet. The variable sq ft is a character variable, so I had to make it numeric, but I had to change some things. First, I removed the commas from the numbers in the dataset to make them ready to be converted to a numeric variable.

```{r}
#| label: Remove commas in all numbers

SoCalRent2025 <- SoCalRent2025 %>% 
  mutate(
    `sq ft` = str_remove(`sq ft`, ",") #Remove commas in all numbers that have them
  )
```

That would make everything ready for converting square feet into a numerical variable, except that some of the entries in `sq ft` are what appear to be ranges with a "-" character. To do this, I ultimately decided to make two new variables: a minimum and a maximum variable for square feet so that both are represented to look at additional nuances of area.

```{r}
#| label: two different datasets that base off the range

SoCalRent2025 <- SoCalRent2025 %>% 
  mutate(
    min_sq_ft = if_else(str_detect(`sq ft`, "-"), 
    true = as.numeric(str_extract(`sq ft`, "[0-9]+")),
    false = as.numeric(`sq ft`),
    missing = NA), # Code to put the minimum square feet of the house, if there is a range of areas
    max_sq_ft = if_else(str_detect(`sq ft`, "-"),
    true = as.numeric(str_remove(str_extract(`sq ft`, "-[0-9]+$"), "-")),
    false = as.numeric(`sq ft`),
    missing = NA) # Code to put the maximum square feet of the house, if there is a range of areas
    )

# If the house has no range, the minimum and maximum are the same.

```

## Exploratory Data Analysis

Let's start with a general summary

```{r}
#| label: exploring the data

SoCalRent2025 |>
  group_by(county) %>% 
  summarize(
    mean_price = mean(price),
    n = n() # Total number of houses by county
  )
  
```

The dataset SoCalRent2025 features three counties: Los Angeles (LA), Orange (OC), and Riverside (RIV). The distribution is fairly even, with each county having 100 houses apiece to ensure everything is almost equally represented. The main thing that stood out was the mean price of LA County and Orange County are both much higher than that of Riverside, though I wanted to focus on LA County and Orange County to see if there was a difference between those two.

```{r}
#| label: seeing the range of prices for each type of house

ggplot(SoCalRent2025, aes(x = price, y = county)) +
  geom_boxplot() +
  labs(
    x = "Price",
    y = "County"
  ) # Boxplots to see the relationship between price and county

 SoCalRent2025 %>%
   group_by(county) %>%
     summarize(
       min_price = min(price),
       q1_price = quantile(price, 0.25),
       med_price = median(price),
       q3_price = quantile(price, 0.75),
       max_price = max(price)
     ) # Numerical summary of the boxplot
```

This graph is displaying the amount of outliers that are present in the dataset in terms of price, and there are two are stick out like a sore thumb.

```{r}
#| label: investigating the outliers in terms of price

SoCalRent2025 %>%
   filter(price >= 20000)
```

Right off the bat, a rent for \$30000 in OC, a house in Anaheim that has 2 bedrooms and 2 bathrooms. It doesn't sound like it is the norm, so we will remove this outlier. We can also filter out the \$20000 rent in LA county because that is the other thing that is way above the rest of the prices, plus it's a West Hills house that has square feet, bedrooms, and bathrooms of all the houses.

```{r}
#| label: removing the massive outliers in terms of Price and Square Feet

SoCalRent2025_1 <- SoCalRent2025 %>% 
  filter(price < 20000) %>% 
  mutate(
    `sq ft` = as.numeric(`sq ft`)
  ) # makes square feet numeric

ggplot(SoCalRent2025_1, aes(x = `sq ft`, y = type)) +
   geom_boxplot()
```

```{r}
#| label: new price by county summary and price five-number summary

SoCalRent2025_1 %>%
  group_by(county) %>%
  summarize(
    mean_price = mean(price)
  )

SoCalRent2025_1 %>%
  summarize(
    min_price = min(price),
    q1_price = quantile(price, 0.25),
    med_price = median(price),
    q3_price = quantile(price, 0.75),
    max_price = max(price)
  )

```

We have the remaining prices, and it appears that Orange County, on average, has higher rent than Los Angeles County. One variable that makes sense to explain is square feet of the houses:

```{r}
#| label: boxplots for square feet by county

ggplot(SoCalRent2025_1, aes(x = min_sq_ft, y = county)) +
   geom_boxplot()

ggplot(SoCalRent2025_1, aes(x = max_sq_ft, y = county)) +
   geom_boxplot()
```

Now, there are two different values due to some having ranges as part of their dataset. Here is a summary of the boxplots:

```{r}
#| label: updated summary of the minimum square feet values

SoCalRent2025_1 %>%
  filter(!is.na(min_sq_ft)) |>
   group_by(county) %>% 
     summarize(
       minimum_sq_ft = min(min_sq_ft),
       q1_sq_ft = quantile(min_sq_ft, 0.25),
       med_sq_ft = median(min_sq_ft),
       q3_sq_ft = quantile(min_sq_ft, 0.75),
       max_sq_ft = max(min_sq_ft),
       mean_sq_ft = mean(min_sq_ft)
     )
```

```{r}
#| label: updated summary of the maximum square feet values

SoCalRent2025_1 %>%
  filter(!is.na(max_sq_ft)) %>% 
   group_by(county) %>%
     summarize(
       min_sq_ft = min(max_sq_ft),
       q1_sq_ft = quantile(max_sq_ft, 0.25),
       med_sq_ft = median(max_sq_ft),
       q3_sq_ft = quantile(max_sq_ft, 0.75),
       maximum_sq_ft = max(max_sq_ft),
       mean_sq_ft = mean(max_sq_ft)
     )
```

So the datasets have quite some notable differences, though it may not be a fair representation as some of the minimum values by county don't even change, labeling things as outliers in maximum prices even though they are not in the minimum prices. Square feet doesn't appear to play a big factor, as they are all extremely similar between the counties.

With everything filtered out in terms of price, here is the final dataset in graph form:

```{r}
#| label: histogram of final dataset by price

condition <- if_else(SoCalRent2025_1$price < 6000, "True", "False")

ggplot(SoCalRent2025_1, aes(x = price, fill = condition)) +
  geom_histogram(bins = 10) +
  labs(
    title = "Rent Distrbution by County",
    x = "Price",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 24,
      face = "bold"
    ),
    plot.margin = margin(t = 10, r = 20, b = 30, l = 10),
    axis.title.x = element_text( # X-axis label
      size = 20
    ),
    axis.title.y = element_text( # X-axis label
      size = 20
    ),
    axis.text.x = element_text(
      size = 12
    ),
    axis.text.y = element_text(
      size = 16
    ),
    panel.spacing.x = unit(20, "points"),
  ) +
  facet_wrap(vars(county), ncol = 3, labeller = labeller(
               county = c(
                 "LA" = "Los Angeles County",
                 "OC" = "Orange County",
                 "RIV" = "Riverside County"
               )
             )) +
  scale_fill_manual(values = c("True" = "red", "False" = "blue"))
```

## Modeling

In order to see the true relationship, let's do some Bayesian Inference.

```{r}
bayes_SoCalRent <- stan_glm(
  price ~ county,
  data = SoCalRent2025_1,
  family = "gaussian",
  prior = normal(location = 4, autoscale = TRUE),
  prior_intercept = default_prior_intercept(),
  seed = 1
)

SoCalRent_dist <- bayes_SoCalRent %>% 
  as.data.frame()

head(SoCalRent_dist)
```

The reference level is LA county, and we are comparing it to Orange County and Riverside County.

```{r}
plot(
  bayes_SoCalRent,
  pars = c("countyOC", "countyRIV"),
  plotfun = "hist",
  bins = 15
) +
  labs(x = "Slope by county")
```

```{r}
predicted_house <- data.frame(
  county = c("OC", "LA", "RIV")
)

predicted_rents <- bayes_SoCalRent %>% 
  posterior_predict(
    newdata = predicted_house,
    draws = 1000,
    seed = 1
  )
```

```{r}
predicted_rents_df <- predicted_rents %>% 
  as.data.frame()

ggplot(data = predicted_rents_df,
       mapping = aes(x = `1`)
       ) +
  geom_histogram(
    bins = 20
  ) +
  labs(title = "Renting a House by county",
       x = "Rent Price",
       y = "Number of Simulated Houses"
  )

bayes_SoCalRent |>
  posterior_interval(
      pars = c("countyOC", "countyRIV"),
      prob = 0.90
    )

```

We can see that there is a 90% chance that the slope of the population model is between -66.89678 and 1097.220. This implies that there is no distinct difference in prices between the counties of LA and OC.
